---
title: "HW3 Attempt 2"
author: "Taylor Thul"
date: "February 14, 2017"
output: html_document
---



You are now the data slave to the principal investigator Dr. Vinca Monster. Dr. M is in the Grape Program at State U, and you are just a poor graduate student trying to get your degree. Dr. M is interested in wine preferences and the influences of physico-chemical properties on preferences. Her laboratory has gathered an extensive dataset on Portugese white varietals.


You will find the white_wines.csv file and its description in my github repo (https://github.com/vhertzb/Regression-1)[https://github.com/vhertzb/Regression-1].

Please use the techniques you have learned in the last two classes, specifically exploratory data analysis and linear regression, to determine association of the wine properties on preference. 

Prepare a report for presentation at the next Monster lab meeting about this dataset.

Rubric:

Exploration (summary statistics (the m's), univariate graphs, multivariate graphs)
Regression (Models explored, diagnostics completed, final model choice, justification)

Please include a concluding paragraph (or two) about the implications of your findings.

```{r echo=FALSE}
# include this code chunk as-is to set options
knitr::opts_chunk$set(comment=NA, prompt=TRUE, out.width=750, fig.height=8, fig.width=8)
library(Rcmdr)
library(car)
library(RcmdrMisc)
```

```{r echo=FALSE}
# include this code chunk as-is to enable 3D graphs
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)

#to get collum names
names(White_wines) <- make.names(names(White_wines))
```

Run some summary statistics: 

```{r}
numSummary(White_wines[,c("alcohol", "chlorides", "citric.acid", "density", 
  "fixed.acidity", "free.sulfur.dioxide", "pH", "quality", "residual.sugar", 
  "sulphates", "total.sulfur.dioxide", "volatile.acidity")], 
  statistics=c("mean", "sd", "IQR", "quantiles"), quantiles=c(0,.25,.5,.75,1))
```

So the main variables we are looking at include alcohol, chlorides, citric acid, density, fixed acidity, free sulfur dioxide, pH, residual sugar, sulphates, total sulfur dixoide, and volatile acidity. We want to know how each of these variables influence the variable 'quality.' 

Make sure no data is missing:

```{r}
sapply(White_wines, function(x)(sum(is.na(x)))) 
# NA counts
```
No missing data! Hooray!

Now to explore each variable indivually. 

Firstly let us look at the variable of interest, quality:
```{r}
with(White_wines, Hist(quality, scale="frequency", breaks="Sturges", col="red"))
#add lables
title (main= "Quality of Wine/ Preference")
```

Based on this histogram the data appears fairly normally distributed.

Since this is the variable of interest we will also use other visual means of exploring the data:

```{r}
Boxplot( ~ quality, data=White_wines, id.method="y")
```

From the boxplot we can see there are a few outliers which will need to be kept in mind during analysis.


Now look at alcohol:
```{r}
with(White_wines, Hist(alcohol, scale="frequency", breaks="Sturges", col="green"))

#add lables
title(main="Histogram of Alcohol % By Volume") 

```

Fairly normal distribution with few outliers. 

Chlorides:
```{r}
with(White_wines, Hist(chlorides, scale="frequency", breaks="Sturges", col="purple"))
title(main="Chlorides in Wine")
```

This appears to be skewed and have some outliers so I will do a boxplot

```{r}
Boxplot( ~ quality, data=White_wines, id.method="y")
```

Citric Acid:
```{r}
with(White_wines, Hist(citric.acid, scale="frequency", breaks="Sturges", col="orange"))
title (main= "Citric Acid in Wine")

scatterplot(quality~citric.acid, reg.line=FALSE, smooth=FALSE, spread=FALSE, boxplots=FALSE, 
  span=0.5, ellipse=FALSE, levels=c(.5, .9), data=White_wines)

Boxplot( ~ citric.acid, data=White_wines, id.method="y")
```

Citric acid also appears to have quite a few outliers and some skewness. 

Density
```{r}
with(White_wines, Hist(density, scale="frequency", breaks="Sturges", col="yellow"))
title(main= "Density in Wine")
```

Density also appears to have a skew with outliers to the right.  I'm noting most of the skewness is to the left so I am now wondering if prehaps these variables are corrolated. I will explore this later after the individual variables.  

Fixed acidity
```{r}
with(White_wines, Hist(fixed.acidity, scale="frequency", breaks="Sturges", col="pink"))
title(main="Fixed Acidity in Wine")

#using a dotplot
with(White_wines, Dotplot(fixed.acidity, bin=FALSE))

Boxplot( ~ fixed.acidity, data=White_wines, id.method="y")
```
Again outliers noted so a box plot reviewed. 

Free sulfur dioxide:
```{r}
with(White_wines, Hist(free.sulfur.dioxide, scale="frequency", breaks="Sturges", col="brown"))
title(main="Free Sulfur Dioxide in Wine")

#using a dotplot
with(White_wines, Dotplot(free.sulfur.dioxide, bin=FALSE))

Boxplot( ~ free.sulfur.dioxide, data=White_wines, id.method="y")

#trying a density plot
densityPlot( ~ free.sulfur.dioxide, data=White_wines, bw="SJ", adjust=1, kernel="gaussian")
```


pH:

```{r}
with(White_wines, Hist(pH, scale="frequency", breaks="Sturges", col="blue"))
title(main="pH level in Wine")


Boxplot( ~ pH, data=White_wines, id.method="y")
```
pH appears to have less outliers and overall more even distribution than the varibles previous.  

Residual sugar:

```{r}
with(White_wines, Hist(residual.sugar, scale="frequency", breaks="Sturges", col="ivory"))
title(main="Residual Sugar in Wine")

Boxplot( ~ residual.sugar, data=White_wines, id.method="y")

```
Continuing the trend of large outliers. 

Sulphates:
I am interested espically in this varible as I have read the level of sulphates influences the taste of wine.  I will see if the data supports that in this case. 

```{r}
with(White_wines, Hist(sulphates, scale="frequency", breaks="Sturges", col="magenta"))
title(main="Free Sulfur Dioxide in Wine")

#using a dotplot
with(White_wines, Dotplot(sulphates, bin=FALSE))

Boxplot( ~ sulphates, data=White_wines, id.method="y")
```
Still some skew and outliers to the right but appears more evenly distributed.  


Total sulfur dixoide:
I anticipate this may be realted to free sulfur dioxide so will keep in mind to test later. 

```{r}
with(White_wines, Hist(total.sulfur.dioxide, scale="frequency", breaks="Sturges", col="limegreen"))
title(main="Total Sulfur Dioxide in Wine")

#using a dotplot
with(White_wines, Dotplot(total.sulfur.dioxide, bin=FALSE))

Boxplot( ~ total.sulfur.dioxide, data=White_wines, id.method="y")
```


Volatile acidity:

```{r}
with(White_wines, Hist(volatile.acidity, scale="frequency", breaks="Sturges", col="lightblue"))
title(main="Volatile Acidity in Wine")


Boxplot( ~ volatile.acidity, data=White_wines, id.method="y")

densityPlot( ~ volatile.acidity, data=White_wines, bw="SJ", adjust=1, kernel="gaussian")
```

The trend continues...

Now I want to look at how these varibles might be related to each other.  

```{r}
install.packages("sm")
library(sm)

sm.density.compare(White_wines$total.sulfur.dioxide, White_wines$volatile.acidity)
#this didn't work with this data set but I want to keep it in the code for another time, they must be in a dataframe to use
```

Compare chlorides, citiric acid, dentsity, and volatile acidity to see any relationship:
```{r}
scatterplotMatrix(~chlorides+citric.acid+density+volatile.acidity, reg.line=FALSE, 
  smooth=TRUE, spread=FALSE, span=0.5, ellipse=FALSE, levels=c(.5, .9), id.n=0, diagonal = 
  'density', data=White_wines)
```

There don't seem to be any assoications standing out from this comparison.  This makes sense because the distribution of the data in all these varibales was quite similar.  

Look at how some variables may be assoicated with the variable of interest, quality. 

```{r}
scatterplotMatrix(~alcohol+chlorides+quality+fixed.acidity+free.sulfur.dioxide, reg.line=FALSE, smooth=TRUE,
   spread=FALSE, span=0.5, ellipse=FALSE, levels=c(.5, .9), id.n=0, diagonal = 'density', 
  data=White_wines)
```
There appear to be some weak associations between quality and fixed acidity as well as free sulfur dioxide.

```{r}
scatterplotMatrix(~quality+citric.acid+density+pH+volatile.acidity, reg.line=FALSE, 
  smooth=TRUE, spread=FALSE, span=0.5, ellipse=FALSE, levels=c(.5, .9), id.n=0, diagonal = 
  'density', data=White_wines)
```
No strong assocaitions stand out except a small negative between qulaity and volatile acidity and a posity between density and volatile acidity.  

One more scatterplot matrix to capture all the variables.

```{r}
scatterplotMatrix(~quality+residual.sugar+sulphates+total.sulfur.dioxide, reg.line=FALSE, 
  smooth=TRUE, spread=FALSE, span=0.5, ellipse=FALSE, levels=c(.5, .9), id.n=0, diagonal = 
  'density', data=White_wines)
```
A slight negative association is seen between quality and total sulfur dioxide as well as a slight positive relationship between sulphates and total sulfar dioxide, which makes sense.  

A quick exploration between varibales with similar names to see any relationship before further analysis

```{r}
scatterplotMatrix(~fixed.acidity+free.sulfur.dioxide+sulphates+total.sulfur.dioxide+volatile.acidity,
   reg.line=FALSE, smooth=TRUE, spread=FALSE, span=0.5, ellipse=FALSE, levels=c(.5, .9), 
  id.n=0, diagonal = 'density', data=White_wines)
```

Some positive association are seen indicating they may be highly related.  

Now lets look at how all thse might go together to predict preference ('quality').
First I will split the data into a training and test data set to ensure the model fits well. 

```{r}
# divide the dataset into a training and a testing set based on a random uniform number on fixed seed, which in this case we are using the date
# this step is also creating a new variable and adding it to the data set which is a distribution of random numbers from 0 to 1 

set.seed(20170214)
White_wines$group <- runif(length(White_wines$quality), min = 0, max = 1)

#what random forests do is this process over and over again and makes the aggregate which might be called bootstrapping?

White_wines.train <- subset(White_wines, group <= 0.90)
White_wines.test <- subset(White_wines, group > 0.90)

#see if it worked
summary(White_wines.train)
summary(White_wines.test)

#I think it worked! 
```

I will use the train data to fit the model and the test data to analysis the fit.

```{r}
RegModel.1 <- 
  lm(quality~alcohol+chlorides+citric.acid+density+fixed.acidity+free.sulfur.dioxide+pH+residual.sugar+sulphates+total.sulfur.dioxide+volatile.acidity,
   data=White_wines.train)
summary(RegModel.1)
```
So it looks like alcohol, chlorides, density, fixed acidity, pH, residual. sugar, sulphates, and volatile acidity are all singificanly assoicated with predicting quality.  
Run the model again taking out the non-signficant variables:

```{r}
RegModel.2 <- 
  lm(quality~alcohol+density+fixed.acidity+free.sulfur.dioxide+pH+residual.sugar+sulphates++volatile.acidity,
   data=White_wines.train)
summary(RegModel.2)
```
In this model the adjusted R-squared is 0.28 (indicating 28% of the variability in average quality rating is due to the varaibles included), exactly the same as the model containing all variables and the F-statistic is highly significant.  

Some diagnositc to ensure fit:
```{r}
residualPlots(RegModel.2)
```


Now there are two additional variables that have a non-zero trend, pH and sulphates.  I will investigate what happens when they are removed from the model.  

```{r}
RegModel.3 <- 
  lm(quality~alcohol+density+fixed.acidity+free.sulfur.dioxide+residual.sugar+volatile.acidity,
   data=White_wines.train)
summary(RegModel.3)
```


In this new model, 27% of the variation is explained by the variables included and of note, fixed acidity does not reach significane.  The F-statistic remains significant, however.  What happens when fixed acidity is removed?

```{r}
RegModel.4 <- 
  lm(quality~alcohol+density+free.sulfur.dioxide+residual.sugar+volatile.acidity,
   data=White_wines.train)
summary(RegModel.4)
```

Now all the variables included are significant, the F-statistic remains significant and the R-squared is only lowered a fraction of a percent.  

Now I will run diagnositics for the updated model.

```{r}
residualPlots(RegModel.4)
```


Now free sulfur dioxide has a non-zero trend.  Can they be removed from the model as well?

```{r}
RegModel.5 <- 
  lm(quality~alcohol+residual.sugar+density+volatile.acidity,
   data=White_wines.train)
summary(RegModel.5)
```

It seems as if it can.  The model without density and free sulfur dioxide has an multiple R-squared of 0.26, slighly lower but nominally so.  The F-statistic and all variables included remain significant.
Now to check the residuals...
```{r}
residualPlots(RegModel.5)
```

This seems to fit well.  Lets continue with assessing fit.

```{r}
#added variable plots
avPlots(RegModel.5, id.n=2, id.cex=0.7)
#id.n - identify n most influential observations
#id.cex - controls the size of the dot

```

```{r}
# run the qq-plot
qqPlot(RegModel.5, id.n=3)
# here, id.n identifies the n observations with the largest residuals in absolute value
```

The distribution may have a slight left skew but is overall normal.  Hooray!

```{r}
#run Bonferroni test for outliers
outlierTest(RegModel.5)
outlierTest(RegModel.4)
```

The Bonferonni p-value remains significant, indicating the model is sound even taking into account the signifant error of using multiple predictor variables. 


```{r}
#identify highly influential points
influenceIndexPlot(RegModel.5, id.n=3)
```
There do appear to be some influential outliers, espcially 2539 based on a quick review.

```{r}
#make influence plot
influencePlot(RegModel.5, id.n=3)
# diameter is related to cooks distance
```
Again observation 2539 appears to be a largly influential outlier.  I would consider taking this point out and seeing if the model still fits.  

```{r}
#attempting to take out point 2539
White_wines.train <- White_wines.train[-c(2539), ]

#re-run model with delted row
RegModel.5 <- 
  lm(quality~alcohol+residual.sugar+density+volatile.acidity,
   data=White_wines.train)
summary(RegModel.5)

#re-run influence plot
influencePlot(RegModel.5, id.n=3)
```

It does appear that removing observation 2539 does make a difference.  Now there are more outliers but the significane of the point is removed without greatly affecting the F-statistic and p-value.  

Now I will use a Breusch–Pagan test for heteroscedasticity.

```{r}
#test for heteroskedasticity
ncvTest(RegModel.5)
```
This indicates (I think) that overall the data has heteroskedasticity. 

Finally I will look at the degree of multicollinearity present amoung variables.  I was initally hypothesising there was at least some multicollinearity between some variables simple based on their names and the distribution of data.  I will first run Model #2 to see how much was present when all significant variables were included in the model.  

```{r}
vif(RegModel.2)
```

Hypothesis confirmed.  There was definatly multicolleniarty issues in the first model.  Now let me look at the final model.  

```{r}
vif(RegModel.5)
```
Better! There is still a sight issues but on the whole they seem to be independent variables.  

So overall the model building process has showing we can use the variables percent alcohol, the amount of residual sugar, wine density and wine volatile acidity to predict the level of quality (or for our purposes preference) that will be given to a certain wine by a drinker.  I chose these specific variables through comparing multiple models for the best fit while maintaining parsimony.

